Open-source tools in R for landscape ecology

### 2021-04-29 paper/paper.tex
Some last typos

While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.}
Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, page 10, Table \ref{tab:packages}).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and for coordinates transformation.
Additionally, \textit{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the higher computational cost \cite{Lovelace2019}.
The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \textit{R} packages directly depend or imports \textit{sp}.
However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
Besides many advantages and strengths of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
Because all \textit{tidyverse} packages follow the same data philosophy structures, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
\textit{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \textit{R} session.
Since publicly available data is becoming more prominent, so are \textit{R} packages to access them.
Packages include \textit{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \textit{elevatr} package to access raster elevation data \cite{Hollister2020}, the \textit{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the \textit{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \textit{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \textit{FedData} package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \textit{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \textit{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \textit{sen2r} \cite{Ranghetti2020} to download Sentinel-2 optical images.
Also, the \texttt{getData()} function from the \textit{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
Additionally, the \textit{rgee} package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig. \ref{fig:fig-survey}a)).

@@

Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \textit{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig. \ref{fig:fig-survey}d)).

>>>>

While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.}
Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, Table \ref{tab:packages}).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and for coordinates transformation.
Additionally, \textit{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the higher computational cost \cite{Lovelace2019}.
The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \textit{R} packages directly depend or imports \textit{sp}.
However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
Besides many advantages and strengths of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
Because all \textit{tidyverse} packages follow the same philosophy how to structure data, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
\textit{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Nowadays, spatial data at various scales is available from many online-accessible sources.
A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \textit{R} session.
Since publicly available data is becoming more prominent, so are \textit{R} packages to access them.
Packages include \textit{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \textit{elevatr} package to access raster elevation data \cite{Hollister2020}, the \textit{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the \textit{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \textit{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \textit{FedData} package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \textit{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \textit{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \textit{sen2r} \cite{Ranghetti2020} to download Sentinel-2 optical images.
Also, the \texttt{getData()} function from the \textit{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
Additionally, the \textit{rgee} package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig. \ref{fig:fig-survey}a).

@@

Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \textit{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig. \ref{fig:fig-survey}d).
